Practical seed-recovery for the PCG Pseudo-Random Number Generator

### 2020-08-28 article/main.tex
dernieresmodifs

% !TeX spellcheck = en
\documentclass[submission,svgnames,journal=tosc]{iacrtrans}

@@

\institute{% ordre alphabétique
Univ. Lille, CNRS, Centrale Lille, UMR 9189 - CRIStAL - Centre de Recherche en Informatique Signal et Automatique de Lille, F-59000 Lille, France \\ 
\email{charles.bouillaguet@univ-lille.fr}
\and 
Sorbonne Université, CNRS, LIP6, F-75005 Paris, France \\
\email{florette.martinez@lip6.fr}
\and 
Sorbonne Université \\
\email{julia.sauvage@etu.upmc.fr}
}

\begin{document}
\maketitle

@@

\begin{abstract}
  The Permuted Congruential Generators (\textsf{PCG}) are popular conventional
  (non-cryptographic) pseudo-random generators designed in 2014. They are used
  by default in the \texttt{NumPy} scientific computing package. Even though
  they are not of cryptographic strength, their designer stated that predicting
  their output should be nevertheless be "challenging".

@@

Pseudo-random generators (PRG) are well-studied primitives in symmetric
cryptography. A PRG is an efficient deterministic algorithm that stretch a small
random seed into a longer pseudo-random stream. To achieve cryptographic-grade
pseudo-randomness, a PRG must ensure that the pseudo-random stream is
computationally indistinguishable from a ``truly'' random sequence of bits by
efficient adversaries. Alternatively, it is possible to define pseudo-randomness
by asking that no efficient algorithm is capable of predicting the next
pseudo-random bit with non-negligible accuracy. The two definitions are in fact
equivalent.

It is well-known that pseudo-random generators can be turned into symmetric
encryption algorithm, by generating ``random'' masks to be used in the one-time
pad. This is precisely what stream ciphers do.

@@

However, problems in non-cryptographic random number generators can also have
dire consequences (barring the obvious case where they are used in lieu of their
cryptographic counterparts). When they are used in scientific computing for
Monte-Carlo methods, their defects have the potential to alter the results of
numerical simulations. Ferrenberg et al.~\cite{Ferrenberg92} ran a classical
Ferromagnetism Ising model Monte-Carlo simulation, in a special case where the
exact results could be computed analytically, and compared the results of the
simulation with the ``true'' answer. They used several conventional
pseudo-random generators: a 32-bit linear congruential generator, two LFSRs,
various combinations thereof, etc. They observed that changing the source of
random numbers significantly altered the outcome of the numerical
simulation. Different generators produced different biases: in particular a
given LFSRs yielded energy levels that were systematically too low and critical
temperatures that were always too high, while another kind of generator yielded
the opposite (in many, repeated, trials).

@@

If $x \in \{0, 1\}^n$ is an $n$-bit string, then $x[i:j]$ denotes the bit string
$x_ix_{i+1}\dots x_{j-2}x_{j-1}$, where $x = x_0\dots x_{n-1}$ (this is the
``slice notation'' used in \textsf{Python}). The set $\mathbb{Z}_{2^k}$ of
integers modulo $2^k$ is seen as the set of $k$-bit strings. If $x$ is a
floating point number, then $\lfloor x \rceil$ denotes the nearest integer
(using the ``rounding half to even'' tie-breaking rule --- this is the default
in IEEE754 arithmetic). If $U$ is a vector or a sequence, then $U_i$ is the
$i$-th element (we use capital letters for these). If $U$ is such a sequence, we
denote by $U \bmod M$ the sequence $(U_0 \bmod M, U_1 \bmod M, \dots)$. The XOR
operation is denoted $\oplus$, left and right rotations are denote $\lll$ and
$\ggg$ respectively. Modular addition is denoted $+$ (or $\boxplus$ to make it
even more explicit).

\textsf{PCG64} has an internal state of 128-bit, which operate as a linear
congruential generator modulo~$2^{128}$. More precisely:
\[
  S_{i+1} = a S_i + c \bmod 2^{128},
\]
Where the ``multiplier'' $a$ is a fixed 126-bit constant. The first initial state $S_0$ is the seed of
the generator. The increment $c$ may be specified by the user of the PRNG to
produce different output streams with the same seed (just as the IV acts in a
stream cipher). If no value of $c$ is specified, then a default increment is
provided. Note that $c$ must be odd. The default values are:
\begin{align*}
  a &= \phantom{1}47026247687942121848144207491837523525 & \text{(fixed)}\\
  c &= 117397592171526113268558934119004209487           & \text{(default value, user-definable)}
\end{align*}

@@

Updating the internal state requires a $128 \times 128 \rightarrow 128$
multiplication operation. In fact, this can be done with three
$64 \times 64 \rightarrow 128$ multiplication and two 64-bit additions. High-end
desktop CPUs all implement these operations in hardware, so the generator is
quite fast on these platforms.

@@

If $M$ is a square matrix, we denote by $\vvvert M \vvvert$ the induced matrix norm :
\[
  \vvvert M \vvvert =  \sup_{x \in \mathbb{R}^n} \frac{\| xM \|}{\| x \|}
\]
In the case of the $\| \cdot \|_2$ norm used throughout this paper, $\| M \|$ is
the largest singular value of $G$; equivalently, it is the square root of the
absolute value of the largest eigenvalue of $M^{t} M$.

@@

When $t$ is greater than the values given in table~\ref{tab:babai}, then
Babai's rounding technique will always return the closest vector, and allow us
to reconstruct a truncated geometric series.

@@

In figure~\ref{fig:Cknown}, we see that guessing $S_0[0:\ell]$ and a few 6-bit
rotations $r_i$ give access to $S_i[58:64+\ell]$ for the corresponding
states. Therefore, looking at $S_i[\ell:64+\ell]$, we are facing a truncated
linear congruential generator on $64$ bits, where we have access to the most
$6+\ell$ bits of each state (denoted by $T$), for a few consecutive states. This
is sufficient to reconstruct entirely the successive states of this truncated
linear congruential generator. This reveals $S_0[\ell:64+\ell]$, and
using~\eqref{eq:output} the entire $S_0$ can be reconstructed. The precise
details follow.

@@

  % S_0
  \begin{scope}
    % remplissage
    \fill[fill=Yellow] (4.6, 0) rectangle +(2.4, 1);
    \fill[fill=magenta] (0, 0) rectangle node {$r_0$} +(0.6, 1);
    \fill[fill=magenta] (11, 0) rectangle node {$w$} +(1.8, 1);

@@

  % S_1
  \begin{scope}[yshift=-3cm]
    % remplissage
    \fill[fill=Yellow] (4.6, 0) rectangle +(2.4, 1);
    \fill[fill=magenta] (0, 0) rectangle node {$r_1$} +(0.6, 1);
    \fill[fill=cyan] (11, 0) rectangle node[text=white] {$w_1$} +(1.8, 1);

@@

  % S_2
  \begin{scope}[yshift=-6cm]
    % remplissage
    \fill[fill=Yellow] (4.6, 0) rectangle +(2.4, 1);
    \fill[fill=magenta] (0, 0) rectangle node {$r_2$} +(0.6, 1);
    \fill[fill=cyan] (11, 0) rectangle node[text=white] {$w_2$} +(1.8, 1);

@@

We will guess $n$ rotations and $\ell$ least-significant bits of the state, for
a total of $2^{6n + \ell}$ guessed bits. Table~\ref{tab:babai} gives a
lower-bound on $\ell$ given $n$, and we see that the total number of guessed
bits reaches a minimum of $38$ when $n=3$ and $\ell=20$. Therefore, success is
guaranteed if we guess $\ell = 20$ low-order bits of the state and three
consecutive rotations.

@@

\item Next, suppose that $\mathcal{C}_i[0:j]$, $S_0[0:j-1]$, $\mathcal{C}_i[64:64+j]$
and $S_0[64:64+j-1]$ are known, for all $i$. We can compute
$\mathcal{C}_i[j] \oplus \mathcal{C}_i[64+j]$ for any $i$ using
eq.~\eqref{eq:master}. We then look a a specific index $\mathfrak{i} > 0$ such that
\[
   \nabla S_{\mathfrak{i}}[j-1] \neq \mathcal{C}_{\mathfrak{i}}[j-1] \quad \text{and} \quad \nabla S_{\mathfrak{i}}[64 + j-1] = \mathcal{C}_{\mathfrak{i}}[64 + j-1].
\]
The point is that, thanks to the majority functions,
$\mathcal{C}_{\mathfrak{i}}[j] = S_0[j-1]$ and
$\mathcal{C}_{\mathfrak{i}}[64+j] = \nabla S_{\mathfrak{i}}[64 + j-1]$. It follows
that:
\[
  S_0[j-1] = Y_0[j-1] \oplus Y_{\mathfrak{i}}[j-1] \oplus \bigl( \nabla S_{\mathfrak{i}}[j-1] \oplus \nabla S_{\mathfrak{i}}[64+j-1] \oplus \nabla S_{\mathfrak{i}}[64 + j-1]\bigr)
\]
From there, we also have $S_0[64+j-1] = Y_0[64+j-1] \oplus S_0[j-1]$, and the
$j$-th carry bits can be computed normally.
\end{itemize}

@@

The analysis of section~\ref{sec:geometric} yields parameters that guarantee
that the reconstruction procedure \emph{always} succeeds. In most cases, these
parameters are pessimistic. We ran a series of experiments to determine more
practical choices: using smaller-than-guaranteed values of $\ell$ (the number of
guessed least-significant bits), we measured the success probability of the
state reconstruction procedure. The results are shown in table~\ref{tab:empirical}.

\begin{table}
  \centering
\begin{tabular}{|c|c|c|c|c|}
  \cline{1-2}\cline{4-5}
  \multicolumn{2}{|c|}{$n=3$ (section~\ref{sec:Cknown})} & &  \multicolumn{2}{c|}{$n=4$ (section~\ref{sec:cvp_small_2})} \\
  \cline{1-2}\cline{4-5}
  $\ell$ & Success proba. && $\ell$ & Success proba. \\
  \cline{1-2}\cline{4-5}
  \noalign{\vskip\doublerulesep
  \vskip-\arrayrulewidth}
  \cline{1-2}\cline{4-5}
  16 & $\approx 0.125$     && 10 &  $\approx 0.12$ \\
  \cline{1-2}\cline{4-5}
  17 & $\approx 0.25$     && 11 &  $\approx 0.64$ \\
  \cline{1-2}\cline{4-5}
  18 & $\approx 0.5$      && 12 &  $\approx 0.995$ \\
  \cline{1-2}\cline{4-5}
  19 & 1                  && 13 &  1 \\
  \cline{1-2}\cline{4-5}
  20 & 1 (proved)         && 14 &  1 (proved) \\
  \cline{1-2}\cline{4-5}
\end{tabular}
\caption{Empirical success probabilities with smaller parameters.}
\label{tab:empirical}
\end{table}

@@

When the increment $c$ is known, algorithm~\ref{algo:known} is all it takes to
reconstruct the internal state of the generator and predict it (or output the
seed). We implemented it in \textsf{C}, using \textsf{OpenMP} to parallelize the
outer loop that guess the least-significant bits of the state. This yields a
simple multi-core implementation. We used the \texttt{gcc 8.3.0} compiler.

>>>>

% !TeX spellcheck = en
\documentclass[journal=tosc,final]{iacrtrans}

@@

\institute{% ordre alphabétique
Univ. Lille, CNRS, Centrale Lille, UMR 9189 - CRIStAL - Centre de Recherche en Informatique Signal et Automatique de Lille, F-59000 Lille, France \\ 
\email{charles.bouillaguet@univ-lille.fr}
\and 
Sorbonne Université, CNRS, LIP6, F-75005 Paris, France \\
\email{florette.martinez@lip6.fr}
\and 
Sorbonne Université, F-75005 Paris, France \\
\email{julia.sauvage@etu.upmc.fr}
}

\begin{document}
\input{settings.tosc}
\maketitle

@@

\begin{abstract}
  The Permuted Congruential Generators (\textsf{PCG}) are popular conventional
  (non-cryptographic) pseudo-random generators designed in 2014. They are used
  by default in the \texttt{NumPy} scientific computing package. Even though
  they are not of cryptographic strength, their designer stated that predicting
  their output should nevertheless be "challenging".

@@

Pseudo-random generators (PRG) are well-studied primitives in symmetric
cryptography. A PRG is an efficient deterministic algorithm that stretches a small
random seed into a longer pseudo-random stream. To achieve cryptographic-grade
pseudo-randomness, a PRG must ensure that the pseudo-random stream is
computationally indistinguishable from a ``truly'' random sequence of bits by
efficient adversaries. Alternatively, it is possible to define pseudo-randomness
by asking that no efficient algorithm is capable of predicting the next
pseudo-random bit with non-negligible accuracy. The two definitions are in fact
equivalent.

It is well-known that pseudo-random generators can be turned into symmetric
encryption algorithms, by generating ``random'' masks to be used in the one-time
pad. This is precisely what stream ciphers do.

@@

However, problems in non-cryptographic random number generators can also have
dire consequences (barring the obvious case where they are used in lieu of their
cryptographic counterparts). When they are used in scientific computing for
Monte-Carlo methods, their defects have the potential to alter the results of
numerical simulations. Ferrenberg et al.~\cite{Ferrenberg92} ran a classical
Ferromagnetism Ising model Monte-Carlo simulation, in a special case where the
exact results could be computed analytically, and compared the results of the
simulation with the ``true'' answer. They used several conventional
pseudo-random generators: a 32-bit linear congruential generator, two LFSRs,
various combinations thereof, etc. They observed that changing the source of
random numbers significantly altered the outcome of the numerical
simulation. Different generators produced different biases: in particular a
given LFSR yielded energy levels that were systematically too low and critical
temperatures that were always too high, while another kind of generator yielded
the opposite (in many, repeated, trials).

@@

If $x \in \{0, 1\}^n$ is an $n$-bit string, then $x[i:j]$ denotes the bit string
$x_ix_{i+1}\dots x_{j-2}x_{j-1}$, where $x = x_0\dots x_{n-1}$ (this is the
``slice notation'' used in \textsf{Python}). The set $\mathbb{Z}_{2^k}$ of
integers modulo $2^k$ is seen as the set of $k$-bit strings. If $x$ is a
floating point number, then $\lfloor x \rceil$ denotes the nearest integer
(using the ``rounding half to even'' tie-breaking rule --- this is the default
in IEEE754 arithmetic). If $U$ is a vector or a sequence, then $U_i$ is the
$i$-th element (we use capital letters for these). If $U$ is such a sequence, we
denote by $U \bmod M$ the sequence $(U_0 \bmod M, U_1 \bmod M, \dots)$. The XOR
operation is denoted $\oplus$, left and right rotations are denoted $\lll$ and
$\ggg$ respectively. Modular addition is denoted $+$ (or $\boxplus$ to make it
even more explicit).

\textsf{PCG64} has an internal state of 128-bit, which operate as a linear
congruential generator modulo~$2^{128}$. More precisely:
\[
  S_{i+1} = a S_i + c \bmod 2^{128},
\]
where the ``multiplier'' $a$ is a fixed 126-bit constant. The first initial state $S_0$ is the seed of
the generator. The increment $c$ may be specified by the user of the PRNG to
produce different output streams with the same seed (just as the IV acts in a
stream cipher). If no value of $c$ is specified, then a default increment is
provided. Note that $c$ must be odd. The default values are:
\begin{align*}
  a &= \phantom{1}47026247687942121848144207491837523525 & \text{(fixed)}\\
  c &= 117397592171526113268558934119004209487           & \text{(default value, user-definable)}
\end{align*}

@@

Updating the internal state requires a $128 \times 128 \rightarrow 128$
multiplication operation. In fact, this can be done with three
$64 \times 64 \rightarrow 128$ multiplications and two 64-bit additions. High-end
desktop CPUs all implement these operations in hardware, so the generator is
quite fast on these platforms.

@@

If $M$ is a square matrix, we denote by $\vvvert M \vvvert$ the induced matrix norm :
\[
  \vvvert M \vvvert =  \sup_{x \in \mathbb{R}^n} \frac{\| xM \|}{\| x \|}
\]
In the case of the $\| \cdot \|_2$ norm used throughout this paper, $\vvvert M \vvvert$ is
the largest singular value of $M$; equivalently, it is the square root of the
absolute value of the largest eigenvalue of $M^{t} M$.

@@

When $t$ is greater than the values given in table~\ref{tab:babai}, then
Babai's rounding technique will always return the closest vector, and allow us
to reconstruct a truncated geometric serie.

@@

In figure~\ref{fig:Cknown}, we see that guessing $S_0[0:\ell]$ and a few 6-bit
rotations $r_i$ give access to $S_i[58:64+\ell]$ for the corresponding
states. Therefore, looking at $S_i[\ell:64+\ell]$, we are facing a truncated
linear congruential generator on $64$ bits, where we have access to the
$6+\ell$ most-significant bits of each state (denoted by $T$), for a few consecutive states. This
is sufficient to reconstruct entirely the successive states of this truncated
linear congruential generator. This reveals $S_0[\ell:64+\ell]$, and
using~\eqref{eq:output} the entire $S_0$ can be reconstructed. The precise
details follow.

@@

  % S_0
  \begin{scope}
    % remplissage
    \fill[fill=yellow] (4.6, 0) rectangle +(2.4, 1);
    \fill[fill=magenta] (0, 0) rectangle node {$r_0$} +(0.6, 1);
    \fill[fill=magenta] (11, 0) rectangle node {$w$} +(1.8, 1);

@@

  % S_1
  \begin{scope}[yshift=-3cm]
    % remplissage
    \fill[fill=yellow] (4.6, 0) rectangle +(2.4, 1);
    \fill[fill=magenta] (0, 0) rectangle node {$r_1$} +(0.6, 1);
    \fill[fill=cyan] (11, 0) rectangle node[text=white] {$w_1$} +(1.8, 1);

@@

  % S_2
  \begin{scope}[yshift=-6cm]
    % remplissage
    \fill[fill=yellow] (4.6, 0) rectangle +(2.4, 1);
    \fill[fill=magenta] (0, 0) rectangle node {$r_2$} +(0.6, 1);
    \fill[fill=cyan] (11, 0) rectangle node[text=white] {$w_2$} +(1.8, 1);

@@

We will guess $n$ rotations and $\ell$ least-significant bits of the state, for
a total of $2^{6n + \ell}$ guessed bits. Table~\ref{tab:babai} gives a
lower-bound on $t=6+\ell$ given $n$, and we see that the total number of guessed
bits reaches a minimum of $38$ when $n=3$ and $\ell=20$. Therefore, success is
guaranteed if we guess $\ell = 20$ low-order bits of the state and three
consecutive rotations.

@@

\item Next, suppose that $\mathcal{C}_i[0:j]$, $S_0[0:j-1]$, $\mathcal{C}_i[64:64+j]$
and $S_0[64:64+j-1]$ are known, for all $i$. We can compute
$\mathcal{C}_i[j] \oplus \mathcal{C}_i[64+j]$ for any $i$ using
eq.~\eqref{eq:master}. We then look a a specific index $\mathfrak{i} > 0$ such that
\[
   \nabla S_{\mathfrak{i}}[j-1] \neq \mathcal{C}_{\mathfrak{i}}[j-1] \quad \text{and} \quad \nabla S_{\mathfrak{i}}[64 + j-1] = \mathcal{C}_{\mathfrak{i}}[64 + j-1].
\]
The point is that, thanks to the majority function,
$\mathcal{C}_{\mathfrak{i}}[j] = S_0[j-1]$ and
$\mathcal{C}_{\mathfrak{i}}[64+j] = \nabla S_{\mathfrak{i}}[64 + j-1]$. It follows
that:
\[
  S_0[j-1] = Y_0[j-1] \oplus Y_{\mathfrak{i}}[j-1] \oplus \bigl( \nabla S_{\mathfrak{i}}[j-1] \oplus \nabla S_{\mathfrak{i}}[64+j-1] \oplus \nabla S_{\mathfrak{i}}[64 + j-1]\bigr)
\]
From there, we also have $S_0[64+j-1] = Y_0[64+j-1] \oplus S_0[j-1]$, and the
$j$-th carry bits can be computed normally.
\end{itemize}

@@

The analysis of section~\ref{sec:geometric} yields parameters that guarantee
that the reconstruction procedure \emph{always} succeeds. In most cases, these
parameters are pessimistic. We ran a serie of experiments to determine more
practical choices: using smaller-than-guaranteed values of $\ell$ (the number of
guessed least-significant bits), we measured the success probability of the
state reconstruction procedure. The results are shown in table~\ref{tab:empirical}.

\begin{table}
  \centering
\begin{tabular}{|c|c|c|c|c|}
  \cline{1-2}\cline{4-5}
  \multicolumn{2}{|c|}{$n=3$ (section~\ref{sec:Cknown})} & &  \multicolumn{2}{c|}{$n=4$ (section~\ref{sec:cvp_small_2})} \\
  \cline{1-2}\cline{4-5}
  $\ell$ & Success proba. && $\ell$ & Success proba. \\
  \cline{1-2}\cline{4-5}
  \noalign{\vskip\doublerulesep
  \vskip-\arrayrulewidth}
  \cline{1-2}\cline{4-5}
  16 & $\approx 0.125$     && 10 &  $\approx 0.12$ \\
  \cline{1-2}\cline{4-5}
  17 & $\approx 0.25$     && 11 &  $\approx 0.64$ \\
  \cline{1-2}\cline{4-5}
  18 & $\approx 0.5$      && 12 &  $\approx 0.995$ \\
  \cline{1-2}\cline{4-5}
  19 & \(\approx 1\)                  && 13 &  \(\approx 1\) \\
  \cline{1-2}\cline{4-5}
  20 & 1 (proved)         && 14 &  1 (proved) \\
  \cline{1-2}\cline{4-5}
\end{tabular}
\caption{Empirical success probabilities with smaller parameters.}
\label{tab:empirical}
\end{table}

@@

When the increment $c$ is known, algorithm~\ref{algo:known} is all it takes to
reconstruct the internal state of the generator and predict it (or output the
seed). We implemented it in \textsf{C}, using \textsf{OpenMP} to parallelize the
outer loop that guesses the least-significant bits of the state. This yields a
simple multi-core implementation. We used the \texttt{gcc 8.3.0} compiler.
